Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-04-04 ms.Rnw
progress

\title[Exact Phylodynamics]{Exact Phylodynamics via Structured Markov Genealogy Processes}
\author[King]{Aaron~A.~King}
\address{
  A.~A.~King,
  Department of Ecology \& Evolutionary~Biology,
  Center for the Study of Complex~Systems, and
  Department of Mathematics,
  University of Michigan,
  Ann~Arbor, MI~48109~USA\\
  Santa~Fe~Institute,
  1399 Hyde~Park~Road,
  Santa~Fe, NM~87501~USA
}
\email{kingaa@umich.edu}
\urladdr{\href{https://kinglab.eeb.lsa.umich.edu/}{https://kinglab.eeb.lsa.umich.edu/}}
\author[Lin]{Qianying~Lin}
\address{
  Q.-Y.~Lin,
  Theoretical Biology and Biophysics,
  Los~Alamos National Laboratory,
  Los~Alamos, NM~87545~USA
}
\author[Ionides]{Edward~L.~Ionides}
\address{
  E.~L.~Ionides,
  Department of Statistics
  University of Michigan,
  Ann~Arbor, MI~48109~USA
}
\date{\today}

@@

@@

We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\mathbb{Z}^d$, but here we assume only that $\Xspace$ is a complete metric measure space with a countable dense subset.
The population process is completely specified by its initial-state density, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}
  \label{eq:ic}
  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
We further assume that $\alpha(t,x,x')$ is \cadlag\ as a function of time for all $x,x'\in{\Xspace}$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
We assume that $\Kr_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+o(\Delta).
  \end{gathered}
\end{equation*}
We will further assume that the number of jumps that occur in a finite time-interval is finite: $\Prob{\Kr_t<\infty}=1$ for all $t$.

\subsubsection{Kolmogorov forward equation}

@@

\AAK{[
    We can also let $\alpha$ have some singular components (i.e., $\delta$ functions).
]}

Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).

@@

\subsubsection{Structured populations, demes, and deme occupancy}
\label{sec:demes}

In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
In particular, we suppose that there are a countable set of subpopulations that may differ in their vital rates, but within each of which, individual lineages are statistically identical.
We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
We define the \emph{deme occupancy} function $n:\Demes\times\Xspace\to\Zp$ so that
for $i\in\Demes$, $x\in\Xspace$, $n_i(x)$ is the number of lineages in deme $i$ when the population is in state $x$.

In the following, it will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
\Cref{fig:markov_state} shows an example with five distinct marks.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

@@

@@

@@

\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
Its state vector is $x=(S,I,R)$ and its KFE is
Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.

@@

A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
The deme occupancy function in this case is $n(x)=(E,I)$.
Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.

@@

@@

@@

Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic, as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\begin{mathsize}{9pt}{10pt}
  \begin{equation}
    \label{eq:Hdens}
    \pi^{\mathrm{H}}(\dd{\H_t})
    =p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}
    \,\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
  \end{equation}
\end{mathsize}
where again, by convention, $\Th^{}_0=0$ and $\Th^{}_{K^{}_t+1}=t$.

@@

\begin{defn}
  Let $\Xr_t$ be a continuous-time Markov process with KFE
  \begin{mathsize}{9pt}{10pt}
    \begin{equation}
      \label{eq:kfe2}
      \frac{\partial{u}}{\partial{t}}(t,x)
      =\int{u(t,x')\,\beta(t,x',x)\,\dd{x'}}
      -\int{u(t,x)\,\beta(t,x,x')\,\dd{x'}}.
    \end{equation}
  \end{mathsize}
  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ and $\lambda:\Rp\times\Xspace\to\mathbb{R}$ are are given measurable functions.
  Let $S\subset\Rp$ be countable and locally finite (\ie $S\cap{[0,t]}$ is finite for all $t>0$).
  Then the system of equations
  \begin{mathsize}{9pt}{11pt}
    \begin{align}
      \frac{\partial{w}}{\partial{t}}(t,x)&=
      \int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -\lambda(t,x)\,w(t,x),
      \qquad
      &t\notin{S},
      \label{eq:filter-eq-defn-reg}\\
      w(t,x)&=\int{\wt(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
      &t\in{S},
      \label{eq:filter-eq-defn-sing}
    \end{align}
    \normalfont
  \end{mathsize}
  is called the \emph{filter equation} \emph{generated by} $\beta$, with \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observation times} $S$.
  The process $\Xr_t$ is said to be the \emph{driver} of the filter equation.
  \Cref{eq:filter-eq-defn-reg} is the \emph{regular part} of the filter equation;
  \Cref{eq:filter-eq-defn-sing} is known as the \emph{singular part}.
\end{defn}

@@

Since there is only a single deme, and since nothing depends on the state, in writing the corresponding filter equation, we can take $w$ to be independent of both $x$ and $y$.
If only $m$ samples are taken simultaneously at a single time, $T$, the filter equation thus reads
\begin{mathsize}{10pt}{11pt}
  \begin{equation}
    \label{eq:moran-filter}
    \begin{gathered}
      w(0) = 1,
      \qquad
      \frac{\partial{w}}{\partial{t}}=
      \mu\,w(t)\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\right)
      -\mu\,w(t),\quad t\notin{S},
      \qquad
      w(t) = \mu\,\frac{\wt(t)}{\tbinom{n}{2}},\quad t\in{S}.
    \end{gathered}
  \end{equation}
\end{mathsize}%
Integrating \cref{eq:moran-filter} and taking logarithms yields
\begin{equation}
  \label{eq:kingman}
  \log{w(T)}=k\,\log{\frac{\mu}{\tbinom{n}{2}}}-\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=m-k}^{m}\!{\tbinom{i}{2}\,s^{}_i},
\end{equation}
where $k$ is the number of branch-points and the $s^{}_{i}\coloneq\int{\Indicator{\ell(t)=i}\,\dd{t}}$ are the durations of the \emph{coalescent intervals}, \ie intervals between successive branch-points.
We recognize \cref{eq:kingman} as the expression for the \citet{Kingman1982a} coalescent \citep[\eg][]{Wakeley2008}.

>>>>

\title[Exact Phylodynamics]{Exact Phylodynamics via Structured Markov Genealogy Processes}
\author[King]{Aaron~A.~King}
\address{
  A.~A.~King,
  Department of Ecology \& Evolutionary~Biology,
  Center for the Study of Complex~Systems, and
  Department of Mathematics,
  University of Michigan,
  Ann~Arbor, MI~48109~USA
  \emph{and} Santa~Fe~Institute,
  1399 Hyde~Park~Road,
  Santa~Fe, NM~87501~USA
}
\email{kingaa@umich.edu}
\urladdr{\href{https://kinglab.eeb.lsa.umich.edu/}{https://kinglab.eeb.lsa.umich.edu/}}
\author[Lin]{Qianying~Lin}
\address{
  Q.-Y.~Lin,
  Theoretical Biology and Biophysics,
  Los~Alamos National Laboratory,
  Los~Alamos, NM~87545~USA
}
\author[Ionides]{Edward~L.~Ionides}
\address{
  E.~L.~Ionides,
  Department of Statistics,
  University of Michigan,
  Ann~Arbor, MI~48109~USA
}
\date{\today}

@@

In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.

@@

We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\mathbb{Z}^d$, but here we assume only that $\Xspace$ is a complete metric measure space with a countable dense subset.
The population process is completely specified by its initial-state density, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}
  \label{eq:ic}
  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
We assume that $\Kr_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+o(\Delta).
  \end{gathered}
\end{equation*}
We further assume that $\alpha(t,x,x')$ is \cadlag\ as a function of time for all $x,x'\in{\Xspace}$ and that the number of jumps that occur in a finite time-interval is finite: $\Prob{\Kr_t<\infty}=1$ for all $t$.

\subsection{Kolmogorov forward equation}

@@

\subsection{Inclusion of jumps at deterministic times}

For modeling purposes, it is sometimes desirable to insist that certain events occur at specific, known times.
For example, if samples are collected at specific times in such a way that the timing itself conveys no information about the process, one might wish to condition on the sampling time.
We can expand the class of population models to allow for this as follows.
Suppose that $S=\Set{s_1,s_2,\dots,}\subset\Zp$ is a sequence of event times.
If we postulate that, at time $s_i$, an event occurred at which $\Xr_t$ jumped from state $x$ to state $x'$ with probability $\pi(s_i,x,x')$, the KFE for the process becomes
\begin{align}
  \label{eq:kfe-reg}
  \frac{\partial{v}}{\partial{t}}(t,x)
  &=\int\!{v(t,x')\,\alpha(t,x',x)\,\dd{x'}}
  -\int\!{v(t,x)\,\alpha(t,x,x')\,\dd{x'}},
  &t\notin{S},\\
  \label{eq:kfe-sing}
  v(t,x)
  &=\int\!{\leftlim{v}(t,x')\,\pi(t,x',x)\,\dd{x'}},
  &t\in{S}.
\end{align}
Note that the \cref{eq:kfe-reg} is identical to \cref{eq:kfe};
we call this the \emph{regular part} of the KFE.
We refer to \cref{eq:kfe-sing} as the \emph{singular part} of the KFE.

Using a Dirac delta notation, one can represent \cref{eq:kfe-reg,eq:kfe-sing} as a single equation in the form of \cref{eq:kfe}.
In particular, if in \cref{eq:kfe} we make the substitution
\begin{equation*}
  \alpha(t,x,x')\mapsto\alpha(t,x,x')+\sum_{s\in{S}}{\delta(t,s)\,\pi(t,x,x')},
\end{equation*}
we obtain an equation which we can view as shorthand for \cref{eq:kfe-reg,eq:kfe-sing}.
%% To be even more parsimonious, we can refer to \cref{eq:kfe-reg,eq:kfe-sing} as the $(\alpha,\pi,S)$ KFE.

%% Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).

@@

It will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
\Cref{fig:markov_state} shows an example with five distinct marks.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

@@

\subsection{Demes and deme occupancy}
\label{sec:demes}

Our first goal in this paper is to show how a given population process induces a unique stochastic process on the space of genealogies.
At each time, this genealogy will represent the relationships of shared ancestry among a population of lineages extant at that time.
To accommodate the structure of the population, this population of lineages will itself be subdivided into discrete categories.
In particular, we suppose that there are a countable set of subpopulations, within each of which individual lineages are statistically identical.
We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
Demes can be used to introduce spatial, behavioral, physiological, and life-history structure into a model.
The flexibility this allows can also be used, via the linear chain trick, to introduce delays into the dynamics.
\Cref{fig:example_models} illustrates this concept in the context of several compartmental models.

We define the \emph{deme occupancy} function $n:\Demes\times\Xspace\to\Zp$ so that
for $i\in\Demes$, $x\in\Xspace$, $n_i(x)$ is the number of lineages in deme $i$ when the population is in state $x$.

@@

The class of population models to which the theory presented here applies is very broad indeed.
In particular, it encompasses the entire class of compartmental models with time-dependent flow rates.
Here, to give a sense of this breadth, we briefly describe a few models of interest.
\Cref{appendix:examples} works out the theory for each of these examples.

@@

\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.

@@

A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
The state space is $\Zp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
It has two demes: $\Demes=\Set{\lab{E},\lab{I}}$.
The deme occupancy function in this case is $n(x)=(E,I)$.
Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.

@@

The linear birth-death process, a mainstay of existing phylodynamic methods, is a special case of the theory presented here.
For this process, we have $\Xspace=\Zp$ and there is a single deme.
$\Xr_t$ represents the size of a population and $n(X_t)=X_t$.

@@

The Kingman coalescent is another workhorse in existing phylodynamic approaches.
It is the ancestral process for the Moran model, in which a fixed population of $n$ lineages experiences events at times distributed according to a rate-$\mu$ Poisson process.
At each such event, an individual lineage selected uniformly at random dies and is replaced by the offspring of a second randomly selected lineage.

@@

Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic, as are $\Kr_t$, the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\begin{mathsize}{9pt}{10pt}
  \begin{equation}
    \label{eq:Hdens}
    \pi^{\mathrm{H}}(\dd{\H_t})
    =p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}
    \,\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
  \end{equation}
\end{mathsize}%
where again, by convention, $\Th^{}_0=0$ and $\Th^{}_{K^{}_t+1}=t$.

@@

\begin{defn}
  Let $\Xr_t$ be a continuous-time Markov process with KFE
  \begin{mathsize}{9pt}{10pt}
    \begin{equation}
      \label{eq:kfe2}
      \frac{\partial{u}}{\partial{t}}(t,x)
      =\int{u(t,x')\,\beta(t,x',x)\,\dd{x'}}
      -\int{u(t,x)\,\beta(t,x,x')\,\dd{x'}}.
    \end{equation}
  \end{mathsize}%
  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ and $\lambda:\Rp\times\Xspace\to\mathbb{R}$ are are given measurable functions.
  Let $S\subset\Rp$ be countable and locally finite (\ie $S\cap{[0,t]}$ is finite for all $t>0$).
  Then the system of equations
  \begin{mathsize}{9pt}{11pt}
    \begin{align}
      \frac{\partial{w}}{\partial{t}}(t,x)&=
      \int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -\lambda(t,x)\,w(t,x),
      \qquad
      &t\notin{S},
      \label{eq:filter-eq-defn-reg}\\
      w(t,x)&=\int{\wt(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
      &t\in{S},
      \label{eq:filter-eq-defn-sing}
    \end{align}
  \end{mathsize}%
  is called the \emph{filter equation} \emph{generated by} $\beta$, with \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observation times} $S$.
  The process $\Xr_t$ is said to be the \emph{driver} of the filter equation.
  \Cref{eq:filter-eq-defn-reg} is the \emph{regular part} of the filter equation;
  \Cref{eq:filter-eq-defn-sing} is known as the \emph{singular part}.
\end{defn}

@@

Since there is only a single deme, and since nothing depends on the state, in writing the corresponding filter equation, we can take $w$ to be independent of both $x$ and $y$.

In the classical \citet{Kingman1982a} case, $m$ samples are taken simultaneously at a single time, $T$.
For this case, if $B$ is the set of branch-times and $\ell(t)$ is the number of lineages in the genealogy at time $t$, the filter equation reads
\begin{mathsize}{10pt}{11pt}
  \begin{equation}
    \label[pluralequation]{eq:moran-filter}
    \begin{gathered}
      w(0) = 1,
      \qquad
      \frac{\partial{w}}{\partial{t}}=
      \mu\,w(t)\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\right)
      -\mu\,w(t),\quad t\notin{B},
      \qquad
      w(t) = \frac{\mu}{\tbinom{n}{2}}\,\wt(t),\quad t\in{B}.
    \end{gathered}
  \end{equation}
\end{mathsize}%
Integrating \cref{eq:moran-filter} and taking logarithms yields
\begin{equation}
  \label{eq:kingman}
  \log{w(T)}=k\,\log{\frac{\mu}{\tbinom{n}{2}}}-\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=m-k}^{m}\!{\tbinom{i}{2}\,s^{}_i},
\end{equation}
where $k=|B|$ is the number of branch-points in $[0,T]$ and the $s^{}_{i}\coloneq\int{\Indicator{\ell(t)=i}\,\dd{t}}$ are the durations of the \emph{coalescent intervals}, \ie intervals between successive branch-points.
We recognize \cref{eq:kingman} as the expression for the \citet{Kingman1982a} coalescent \citep[\eg][]{Wakeley2008}.

More generally, if in addition samples are taken through time at a per-capita rate $\psi(t)$ such that the set of sample-times in the genealogy is $S=S_0\cup{S_1}$, where $S_0$, $S_1$ are the sets of times of terminal and inline samples, respectively, then the filter equation reads
\begin{mathsize}{10pt}{11pt}
  \begin{equation}
    \label[pluralequation]{eq:moran-filter2}
    \begin{gathered}
      w(0) = 1,
      \qquad
      \frac{\partial{w}}{\partial{t}}=
      -\mu\,\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\,w(t),
      \quad t\notin{S\cup{B}},
      \qquad
      w(t) = \frac{\mu}{\tbinom{n}{2}}\,\wt(t),\quad t\in{B},\\[8pt]
      w(t) = \psi(t)\,\left(1-\frac{\ell(t)}{n}\right)\,\wt(t),\quad t\in{S_0},
      \qquad
      w(t) = \frac{\psi(t)}{n}\,\wt(t),\quad t\in{S_1}.
    \end{gathered}
  \end{equation}
\end{mathsize}%
Integrating \cref{eq:moran-filter2} yields
\begin{equation}
  \label{eq:mgp}
  \log{w(T)}=
  |B|\,\log{\frac{\mu}{\tbinom{n}{2}}}
  +\sum_{t\in{S_0}}\log{\left[\psi(t)\,\left(1-\frac{\ell(t)}{n}\right)\right]}
  +\sum_{t\in{S_1}}\log{\left[\frac{\psi(t)}{n}\right]}
  -\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=1}^{\infty}\!{\tbinom{i}{2}\,s^{}_i}.
\end{equation}
